---
title: "Chicago GTFS data ingestion"
author: "jan.zymla"
output: html_notebook
---

```{r}
library(tidyverse)
library(data.table)
library(magrittr)
library(lubridate)
library(purrr)
```
GTFS data feed is available at `https://www.transitchicago.com/downloads/sch_data`
```{r}
library(uuid)
library(reticulate)
library(docker)
library(RPostgres)
```
```{r}
getwd()
```

```{r}
client <- docker$from_env()
```

```{r}
data_folder_path = normalizePath('data_raw', winslash = '\\')
vols <- reticulate::dict()
vols[data_folder_path] <- reticulate::dict(bind = '/extdata/', mode = 'ro')
vols
```
```{r}
ports <- reticulate::dict('5432/tcp' = reticulate::tuple('127.0.0.1', as.integer(1100)))
ports
```
```{r}
pwd <- uuid::UUIDgenerate() %>% as.character()
env <- reticulate::dict('POSTGRES_PASSWORD' = pwd)
env
```

## Start the container
```{r}
pg <- client$containers$run(name = 'pg', image = "starefossen/pgrouting:latest", remove = TRUE, ports = ports, volumes = vols, detach = TRUE, environment = env)
```

# Download and extract GTFS data
```{r}
temp <- tempfile()
download.file('https://www.transitchicago.com/downloads/sch_data/google_transit.zip', temp)
```
```{r}
unzip(zipfile = temp, exdir = data_folder_path)
```



# Now let's connect to the database
```{r}
pg_host <- 'localhost';
pg_port <- '1100';
pg_database <- 'postgres';
pg_user <- 'postgres';
pg_password <- pwd
```


```{r}
con <- 
  dbConnect(
    Postgres(),
    dbname = pg_database,
    host = pg_host,
    port = pg_port,
    user = pg_user,
    password = pg_password
    )
```


```{sql, connection=con}
DROP TABLE IF EXISTS agency;
```


```{sql, connection=con}
CREATE TABLE agency
(
--  agency_id VARCHAR(100) NOT NULL PRIMARY KEY,
  agency_name VARCHAR(100),
  agency_url TEXT,
  agency_timezone VARCHAR(100),
  agency_lang VARCHAR(50),
  agency_phone VARCHAR(50),
  agency_fare_url VARCHAR(100)
)
```

```{sql, connection=con}
COPY agency(agency_name, agency_url, agency_timezone, agency_lang, agency_phone, agency_fare_url) 
FROM '/extdata/agency.txt' DELIMITER ',' CSV HEADER;
```
```{sql connection=con}
CREATE TABLE calendar
(
  service_id VARCHAR(100) NOT NULL,
  monday SMALLINT,
  tuesday SMALLINT,
  wednesday SMALLINT,
  thursday SMALLINT,
  friday SMALLINT,
  saturday SMALLINT,
  sunday SMALLINT,
  start_date DATE,
  end_date DATE,
  start_date_str CHAR(8),
  end_date_str CHAR(8));
```
```{sql, connection=con}
COPY calendar(service_id, monday, tuesday, wednesday, thursday, friday, saturday, sunday, start_date_str, end_date_str) 
FROM '/extdata/calendar.txt' DELIMITER ',' CSV HEADER;
```

```{sql connection=con}
UPDATE calendar SET start_date = to_date(start_date_str, 'YYYYMMDD')::date;
```
```{sql connection=con}
UPDATE calendar SET end_date = to_date(end_date_str, 'YYYYMMDD')::date;
```
```{sql connection = con}
CREATE TABLE calendar_dates
(
  service_id VARCHAR(100) NOT NULL,
  date DATE,
  exception_type SMALLINT,
  date_str CHAR(8)
  );
```

```{sql connection = con}
COPY calendar_dates(service_id, date_str, exception_type) 
FROM '/extdata/calendar_dates.txt' DELIMITER ',' CSV HEADER;
```

```{sql connection = con}
UPDATE calendar_dates SET date = to_date(date_str, 'YYYYMMDD')::date;
```
```{sql connection = con}
DROP TABLE IF EXISTS stop_times;
```
```{sql connection = con}
CREATE TABLE stop_times
(
  trip_id VARCHAR(100) NOT NULL,
  arrival_time interval,
  departure_time interval,
  stop_id VARCHAR(100),
  stop_sequence INT,
  stop_headsign VARCHAR(50),
  pickup_type SMALLINT,
  shape_dist_traveled INT);
```
```{sql connection = con}
COPY stop_times(trip_id,arrival_time,departure_time,stop_id,stop_sequence,stop_headsign,pickup_type,shape_dist_traveled) 
FROM '/extdata/stop_times.txt' DELIMITER ',' CSV HEADER;
```
```{sql connection = con}
CREATE TABLE routes
(
  route_id VARCHAR(100) NOT NULL PRIMARY KEY,
-- agency_id VARCHAR(100),
  route_short_name VARCHAR(100),
  route_long_name VARCHAR(100),
--  route_desc VARCHAR(100),
  route_type SMALLINT,
  route_url VARCHAR(100),
  route_color VARCHAR(100),
  route_text_color VARCHAR(100));
```
```{sql connection = con}
COPY routes(route_id, route_short_name, route_long_name, route_type, route_url, route_color, route_text_color) 
FROM '/extdata/routes.txt' DELIMITER ',' CSV HEADER;
```
```{sql connection = con}
DROP TABLE IF EXISTS stops;
```
```{sql connection = con}
CREATE TABLE stops
(
  stop_id VARCHAR(100) NOT NULL PRIMARY KEY,
  stop_code VARCHAR(100),
  stop_name VARCHAR(100),
  stop_desc VARCHAR(100),
  stop_lat NUMERIC,
  stop_long NUMERIC,
--  zone_id VARCHAR(100),
--  stop_url VARCHAR(100),
  location_type SMALLINT,
  parent_station VARCHAR(100),
  wheelchair_boarding VARCHAR(100));
```
```{sql connection = con}
COPY stops(stop_id, stop_code, stop_name, stop_desc, stop_lat, stop_long, location_type, parent_station, wheelchair_boarding) 
FROM '/extdata/stops.txt' DELIMITER ',' CSV HEADER;
```
```{sql connection = con}
DROP TABLE IF EXISTS transfers;
```
```{sql connection = con}
CREATE TABLE transfers
(
  from_stop_id VARCHAR(100) NOT NULL,
  to_stop_id VARCHAR(100) NOT NULL,
  transfer_type SMALLINT,
--  min_transfer_time  SMALLINT,
  PRIMARY KEY(from_stop_id, to_stop_id)
);
```
```{sql connection = con}
COPY transfers(from_stop_id, to_stop_id, transfer_type) 
FROM '/extdata/transfers.txt' DELIMITER ',' CSV HEADER;
```
```{sql connection = con}
CREATE FUNCTION seconds_to_interval(int) RETURNS interval
    AS $$ SELECT make_interval(secs => $1) $$
    LANGUAGE SQL;
```
```{sql connection = con}
ALTER TABLE transfers ADD COLUMN min_transfer_time interval;
```
```{sql connection = con}
UPDATE transfers SET min_transfer_time = seconds_to_interval(60);
```
```{sql connection = con}
DROP TABLE IF EXISTS trips;
```
```{sql connection = con}
CREATE TABLE trips(
  route_id VARCHAR(100),
  service_id VARCHAR(100),
  trip_id  VARCHAR(100) NOT NULL PRIMARY KEY,
--  trip_headsign  VARCHAR(100),
--  trip_short_name VARCHAR(100),
  direction_id SMALLINT,
  block_id VARCHAR(100),
  shape_id VARCHAR(100),
  direction VARCHAR(50),
  wheelchair_accessible SMALLINT,
--  bikes_allowed SMALLINT,
--  trip_desc VARCHAR(100),
  schd_trip_id  VARCHAR(100)
  );
```
```{sql connection = con}
COPY trips(route_id,service_id,trip_id,direction_id,block_id,shape_id,direction,wheelchair_accessible,schd_trip_id) 
FROM '/extdata/trips.txt' DELIMITER ',' CSV HEADER;
```

```{python}
import pandas as pd
route_type_names = pd.DataFrame(data={'route_type': [0, 1, 2, 3, 7], 'route_type_name': ['tram', 'metro', 'rail', 'bus', 'funicular']})
route_type_names.reset_index().drop('index', 1)
route_type_names
```
```{r}
route_type_names <- 
  tribble(
    ~route_type, ~route_type_name,
    0, 'tram',
    1, 'metro',
    2, 'rail',
    3, 'bus',
    7, 'funicular'
  )
route_type_names
```
```{r}
dbWriteTable(
  con, 
  'route_type_names',
  route_type_names, 
  overwrite = TRUE, 
  field.types = c('route_type' = 'SMALLINT', 'route_type_name' = 'VARCHAR(100)')
  )
```


# GTFS data processing
```{sql connection = con}
CREATE TABLE routes_rail AS
SELECT routes.*, route_type_name FROM
routes JOIN route_type_names ON routes.route_type = route_type_names.route_type
WHERE route_type_names.route_type_name != 'bus';
```
## calender_dates
exception_type	Required	The exception_type indicates whether service is available on the date specified in the date field.
* A value of 1 indicates that service has been added for the specified date.
* A value of 2 indicates that service has been removed for the specified date.
For example, suppose a route has one set of trips available on holidays and another set of trips available on all other days. You could have one service_id that corresponds to the regular service schedule and another service_id that corresponds to the holiday schedule. For a particular holiday, you would use the calendar_dates.txt file to add the holiday to the holiday service_id and to remove the holiday from the regular service_id schedule.
```{sql connection = con}
DROP TABLE IF EXISTS trips_rail_tuesday;
```
```{sql connection = con}
CREATE TABLE trips_rail_tuesday AS
SELECT trips_rail.* FROM
(
	SELECT
		trips.*,
		route_short_name,
		route_long_name,
		route_color,
		route_text_color,
        route_type_name
	FROM trips
	INNER JOIN routes_rail
	ON trips.route_id = routes_rail.route_id
 ) trips_rail INNER JOIN 
(SELECT coalesce(calendar.service_id, calendar_dates.service_id) AS service_id -- date(date_trunc('week', date '2018-11-18' + 5)) + 1, calendar.service_id, calendar_dates.service_id, *
FROM calendar FULL JOIN calendar_dates
ON calendar.service_id = calendar_dates.service_id AND date(date_trunc('week', date '2018-11-12' + 5)) + 1 = calendar_dates.date
 WHERE	(date(date_trunc('week', date '2018-11-18' + 5)) + 1 BETWEEN start_date AND end_date + 1
 			AND tuesday = 1
			AND (exception_type IS NULL OR exception_type = 1)
		) 
	OR (exception_type = 1 AND date = date(date_trunc('week', date '2018-11-18' + 5)))) services 
ON trips_rail.service_id = services.service_id;
```
```{sql connection = con}
DROP TABLE IF EXISTS stop_times_rail_tuesday;
```
```{sql connection = con}
CREATE TABLE stop_times_rail_tuesday AS
SELECT 
    stops.stop_id,
    stop_name,
    stop_lat,
    stop_long,
	trips_rail_tuesday.*,
	arrival_time,
	departure_time,
	stop_sequence,
	pickup_type -- ,
--	drop_off_type
FROM trips_rail_tuesday
INNER JOIN stop_times ON trips_rail_tuesday.trip_id = stop_times.trip_id
LEFT JOIN stops ON stops.stop_id = stop_times.stop_id;
```
```{sql connection = con}
DROP TABLE IF EXISTS nodes_gtfs;
```
```{sql connection = con}
CREATE TABLE nodes_gtfs AS
SELECT
	trip_id||'::'||stop_sequence||'::dep' AS node_id,
	departure_time||' ('||route_short_name||') '||stop_name||' ->' AS node_name,
	'stop_departure' AS node_type,
	departure_time AS node_time,
	stop_id,
	stop_name,
	stop_lat AS lat,
	stop_long AS lon,
	route_id,
--	route_short_name,
--	route_long_name,
    route_type_name,
	route_color,
	route_text_color,
    trip_id,
    stop_sequence -- ,
--	trip_short_name,
--	trip_headsign
	FROM stop_times_rail_tuesday
UNION ALL
SELECT
	trip_id||'::'||stop_sequence||'::arr' AS node_id,
	arrival_time||' ('||route_short_name||') '||stop_name||' <-' AS node_name,
	'stop_arrival' AS node_type,
	arrival_time AS node_time,
	stop_id,
	stop_name,
	stop_lat AS lat,
	stop_long AS lon,
	route_id,
--	route_short_name,
--	route_long_name,
    route_type_name,
	route_color,
	route_text_color,
    trip_id,
    stop_sequence --,
--	trip_short_name,
--	trip_headsign
	FROM stop_times_rail_tuesday
UNION ALL
SELECT DISTINCT ON (stop_id)
	stop_id||'::entry' AS node_id,
	'entry_node : ' || stop_name AS node_name,
	'entry_node' AS node_type,
	CAST(NULL AS INTERVAL) AS node_time,
	stop_id,
	stop_name,
	stop_lat AS lat,
	stop_long AS lon,
	CAST(NULL AS TEXT) AS route_id,
--	CAST(NULL AS TEXT) AS route_short_name,
--	CAST(NULL AS TEXT) AS route_long_name,
    'walk' AS route_type_name,
	CAST(NULL AS TEXT) AS route_color,
	CAST(NULL AS TEXT) AS route_text_color,
    CAST(NULL AS TEXT) AS trip_id,
    CAST(NULL AS SMALLINT) AS stop_sequence --,
--	CAST(NULL AS TEXT) AS trip_short_name,
--	CAST(NULL AS TEXT) AS trip_headsign
	FROM stop_times_rail_tuesday
UNION ALL
SELECT
	stop_id||'::exit' AS node_id,
	'exit_node : ' || stop_name AS node_name,
	'exit_node' AS node_type,
	CAST(NULL AS INTERVAL) AS node_time,
	stop_id,
	stop_name,
	stop_lat AS lat,
	stop_long AS lon,
	CAST(NULL AS TEXT) AS route_id,
--	CAST(NULL AS TEXT) AS route_short_name,
--	CAST(NULL AS TEXT) AS route_long_name,
    'walk' AS route_type_name,
	CAST(NULL AS TEXT) AS route_color,
	CAST(NULL AS TEXT) AS route_text_color,
    CAST(NULL AS TEXT) AS trip_id,
    CAST(NULL AS SMALLINT) AS stop_sequence -- ,
--	CAST(NULL AS TEXT) AS trip_short_name,
--	CAST(NULL AS TEXT) AS trip_headsign
	FROM stop_times_rail_tuesday
;
```
```{sql connection = con}
DROP TABLE IF EXISTS edges_gtfs;
```
```{sql connection = con}
CREATE TABLE edges_gtfs AS
SELECT
	'wait_aboard' AS edge_type,
	nodes_stop_arrivals.node_id AS from_node_id,
	nodes_stop_arrivals.node_name AS from_node_name,
	nodes_stop_arrivals.node_time AS from_node_time,
	nodes_stop_departures.node_id AS to_node_id,
	nodes_stop_departures.node_name AS to_node_name,
	nodes_stop_departures.node_time AS to_node_time,
	EXTRACT(EPOCH FROM (nodes_stop_departures.node_time - nodes_stop_arrivals.node_time)) AS cost
FROM (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_departure'
        ) nodes_stop_departures
JOIN (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_arrival'
        ) nodes_stop_arrivals
ON nodes_stop_departures.trip_id = nodes_stop_arrivals.trip_id
AND nodes_stop_departures.stop_sequence = nodes_stop_arrivals.stop_sequence

UNION ALL

SELECT
	nodes_stop_departures.route_type_name AS edge_type,
	nodes_stop_departures.node_id AS from_node_id,
	nodes_stop_departures.node_name AS from_node_name,
	nodes_stop_departures.node_time AS from_node_time,
	nodes_stop_arrivals.node_id AS to_node_id,
	nodes_stop_arrivals.node_name AS to_node_name,
	nodes_stop_arrivals.node_time AS to_node_time,
	EXTRACT(EPOCH FROM (nodes_stop_arrivals.node_time - nodes_stop_departures.node_time)) AS cost
FROM (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_departure'
        ) nodes_stop_departures
INNER JOIN (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_arrival'
        ) nodes_stop_arrivals
ON nodes_stop_departures.trip_id = nodes_stop_arrivals.trip_id
AND nodes_stop_departures.stop_sequence + 1 = nodes_stop_arrivals.stop_sequence

UNION ALL

SELECT
	'transfer_same_stop' AS edge_type,
	nodes_stop_arrivals.node_id AS from_node_id,
	nodes_stop_arrivals.node_name AS from_node_name,
	nodes_stop_arrivals.node_time AS from_node_time,
	nodes_stop_departures.node_id AS to_node_id,
	nodes_stop_departures.node_name AS to_node_name,
	nodes_stop_departures.node_time AS to_node_time,
	EXTRACT(EPOCH FROM (nodes_stop_departures.node_time - nodes_stop_arrivals.node_time)) AS cost
FROM (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_departure'
        ) nodes_stop_departures
INNER JOIN (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_arrival'
        ) nodes_stop_arrivals
    ON nodes_stop_departures.stop_id = nodes_stop_arrivals.stop_id
	AND nodes_stop_departures.node_time >= nodes_stop_arrivals.node_time + '30 secs'
	AND nodes_stop_departures.node_time <= nodes_stop_arrivals.node_time + '1 hour'
    
UNION ALL
SELECT
	'transfer' AS edge_type,
	nodes_stop_arrivals.node_id AS from_node_id,
	nodes_stop_arrivals.node_name AS from_node_name,
	nodes_stop_arrivals.node_time AS from_node_time,
	nodes_stop_departures.node_id AS to_node_id,
	nodes_stop_departures.node_name AS to_node_name,
	nodes_stop_departures.node_time AS to_node_time,
	EXTRACT(EPOCH FROM (nodes_stop_departures.node_time - nodes_stop_arrivals.node_time)) AS cost
FROM (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_arrival'
        ) nodes_stop_arrivals 
    JOIN transfers ON nodes_stop_arrivals.stop_id = transfers.from_stop_id
    JOIN (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_departure'
        ) nodes_stop_departures
        ON nodes_stop_departures.stop_id = transfers.to_stop_id
WHERE nodes_stop_departures.node_time >= nodes_stop_arrivals.node_time + min_transfer_time
	AND nodes_stop_departures.node_time <= nodes_stop_arrivals.node_time + '1 hour'

UNION ALL

SELECT
	'entry_edge' AS edge_type,
	nodes_entry.node_id AS from_node_id,
	nodes_entry.node_name AS from_node_name,
	nodes_entry.node_time AS from_node_time,
	nodes_stop_departures.node_id AS to_node_id,
	nodes_stop_departures.node_name AS to_node_name,
	nodes_stop_departures.node_time AS to_node_time,
	0::INTEGER AS cost
FROM (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_departure'
        ) nodes_stop_departures
JOIN (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'entry_node'
        ) nodes_entry ON nodes_stop_departures.stop_id = nodes_entry.stop_id
UNION ALL
SELECT
	'exit_edge' AS edge_type,
	nodes_stop_arrivals.node_id AS from_node_id,
	nodes_stop_arrivals.node_name AS from_node_name,
	nodes_stop_arrivals.node_time AS from_node_time,
	nodes_exit.node_id AS to_node_id,
	nodes_exit.node_name AS to_node_name,
	nodes_exit.node_time AS to_node_time,
	0::INTEGER AS cost
FROM (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'stop_arrival'
        ) nodes_stop_arrivals 
JOIN (
        SELECT * 
        FROM nodes_gtfs
        WHERE node_type = 'exit_node'
        ) nodes_exit ON nodes_stop_arrivals.stop_id = nodes_exit.stop_id


```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
```{sql connection = con}

```
